# 5: Dimensionality reduction techniques


Downloading the data, and showing summaries and graphical overview
```{r}
library(dplyr)
library(GGally)
dat<-read.csv("./data/human2.csv")
row.names(dat)<-dat$X
dat<-dat[,2:9]
summary(dat)
ggpairs(dat)
```
Data has 155 observations and 8 variables. 

VARIABLES:
country= Country
liexp= life expectancy
gni= Gross National Income per capita
liexp = Life expectancy at birth
edu = Expected years of schooling 
mat_mor = Maternal mortality ratio
abr = Adolescent birth rate
repre_parl = Percetange of female representatives in parliament
sedu_rat = ratio of females with at least secondary education divided by males with at least secondary education.
lab_rat= proportion of females in work force divided by the proportion of males in work force.

Most of the variables seem to normally distributed. abr seems to be skewed to the right, as is mat_mor. gni is heavily skewed to the right. 
Life-expectancy seems. 


Principal component analysis (PCA) on the raw (non-standardized) human data and bi-plot
```{r}
A<-prcomp(dat)
summary(A)

s<-summary(A)
pca_pr <- round(1*s$importance[2, ], digits = 5)
pca_pr2<-round(pca_pr*100,digits = 2)
pca_lab<-paste0(names(pca_pr), " (", pca_pr2, "%)")

biplot(A, choices = 1:2, cex = c(0.8, 1), col = c("grey70", "deeppink2"), xlab = pca_lab[1], ylab = pca_lab[2])
```

PCA on standardized data
```{r}
dat1<-scale(dat)
dat1<-as.data.frame(dat1)
B<-prcomp(dat1)
summary(B)

s<-summary(B)
pca_pr <- round(1*s$importance[2, ], digits = 5)
pca_pr2<-round(pca_pr*100,digits = 2)
pca_lab<-paste0(names(pca_pr), " (", pca_pr2, "%)")

biplot(B, choices = 1:2, cex = c(0.8, 1), col = c("grey70", "deeppink2"), xlab = pca_lab[1], ylab = pca_lab[2])
```

Both plot and the explained variance changed drastically. PCA seeks to explain variance. Without standardizing, the first variable seems to explain most of the variance. After standardizing it is obvious that other components also contribute, but only the first two components explain more than 10% of the variance. 

In the first plot, only gni /which happens to be the first variable) seems to explain the variance, whereas in the second plot maternal mortality, birth rate, education, life expectancy and ratio of the secondary education seem to explain the first component and ratio of labour force and representation of women in parliament explain the second component.


Interpretations of the obtained results

It seems that first component relates to welfare statish and the component2 more to equity of the sexes.  When looking at the countries in the plot, most of the western states are located in the left, whereas less wealthy countires are located in the right. When looking at the second component, countries in which the status of women is not equal to that of men are located at the bottom of plot. INterestingly some of the african countries are at the top of the plot. It is worth noting that there may be some interactions between the components chosen by pca, as in wealthier countries it may possible for the other sex to not work, whereas in poorer countries both sexes may  have to work

#############
The tea data comes from the FactoMineR package and it is measured with a questionnaire on tea: 300 individuals were asked how they drink tea (18 questions) and what are their product's perception (12 questions). In addition, some personal details were asked (4 questions).

Load the tea dataset and convert its character variables to factors:

```{r}
tea <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/tea.csv", stringsAsFactors = TRUE)


library(dplyr)
library(tidyr)
str(tea)
dim(tea)
View(tea)

tea1<-(tea[,1:10])
ggpairs(tea[,1:10])

pivot_longer(tea1, cols = everything()) %>% 
  ggplot(aes(value)) + facet_wrap("name", scales = "free") + geom_bar()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```
Exploring the data briefly. 

```{r}
library(FactoMineR)

mca <- MCA(tea1, graph = FALSE)

# summary of the model
summary(mca)

# visualize MCA
plot(mca, invisible=c("ind"), graph.type = "classic",habillage = "quali")


```

According the plot, having tea at diner or not at home have the greatest impact on the first 2 dimensions, variables close to horizontal x-axis have no lesser effect on dimension 2 and variables closer to y axis have less effect on dimension1. Most of the "not" variables are close to origo, and their counterparts are located further away, meaning that they have impact on the two dimensions.

Dimension 2 seems to reflect the social affect of drinking tea. Factors related to social aspects of drinking tea seem to affect the y-value in the plot. Dimension 1 is harder to interpret, but one possibility is that dim1 reflects the "tea as a hobby" kind of aspect. things  like  drinking tea in a tearoom and having tea time.



